---
title: "Regression Models Course Project"
author: "kimnewzealand"
date: "28 June 2017"
output:
  pdf_document: default
  latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev='pdf',echo = TRUE)
```

**Load packages**

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
options(digits=3)
devtools::install_github("rstudio/rmarkdown")
library(car)
```

##1. Executive Summary

Motor Trend are interested in exploring the relationship between a set of variables and miles per gallon (MPG). Using a data set of a collection of cars, we take a look at answering the following questions:  
- Is an automatic or manual transmission better for MPG?  
- Quantify the MPG difference between automatic and manual transmissions.

We will perform some EDA then fit three models, a linear model and two multivariable linear models and show that the third one using model selection may be a better model fit based on adjusted R-squared.

##2. Load data

```{r load-data}
data(mtcars)
```

##3. Perform basic exploratory data analysis

```{r headsummary, }
## Create a summary of the top 2 records from mtcars dataset
kable(head(mtcars,2), caption="Summary of first rows of mtcars Dataset",align = c("c", "c"))
```
  
The mtcars data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles. Each row in the dataset is a make of car, with each column as a aspect of design and performance. 

From the violin plot (Figure 1 in the Appendix) it appears that the _mpg_ for the manual transmissions have greater _mpg_ than for the automatic transmissions. See also the summary statistics (Table 2).

We will take a look at fitting different models based on a hypothesis test.


##4. Regression Models 

**HYPOTHESIS TEST**

First set the hypothesis test for the question "Is an automatic or manual transmission better for MPG?":  

The null hypothesis $H_0 : \beta_1 = 0$ the manual transmission is not a significant predictor for _mpg_.  
The alternative hypothesis $H_A : \beta_1 \neq 0$ is that manual transmission is a significant predictor for _mpg_.

We assume for the test that the sampled car types are independent of each other.

**SIMPLE LINEAR REGRESSION MODEL**

The first model we will apply is the simple linear model using the lm function in R on the factor of the categorical predictor variable _am_ with levels automatic transmissions (0) and manual transmissions (1), of the numerical response variable _mpg_.
```{r linearmodel lm1}
##  Create a linear regression model lm1
lm1 <- lm(mpg~factor(am),mtcars)
```

From the coefficient summary in the Appendix, the p-value is < 5%, therefore we would reject the null hypothesis in favour of the alternative hypothesis that the manual transmission is a significant predictor given no other variables are present in the model.

The adjusted R squared is `r round(summary(lm1)$adj.r.squared,2)` which is not very high so this may not be the best model yet.
There may be other variables that impact _mpg_ so we will investigate with a multivariable linear model.

**MULTIVARIABLE LINEAR REGRESSION MODEL**

The second model we will apply is the multivariable linear model, to view if the transmission type (_am_) is a significant predictor, when other significant variables are included in the model.
```{r MVR}
# Create a multivariable linear model of mpg to all the other 10 variables
lm2 <- lm(mpg~.,mtcars)
```
The p-values for the included variables  are 
`r round(summary(lm2)$coefficients[,4],2)` which are all greater than 0.05%, so we would not reject the null hypothesis, given all other variables included in the model. 

We will apply stepwise backward model selection in a third model.

```{r stepwise}
# Calculate model using stepwise backwards model selection.
sw <- step(object = lm2,direction = "backward",trace =FALSE)
```
The adjusted R squared for this third model is `r round(summary(sw)$adj.r.squared,2)`.

We can quantify and interpret the third model further by saying that the manual transmission appears to be a significant predictor of _mpg_ and we may expect an increase of `r round(summary(sw)$coefficients[4,1],2)` _mpg_ when choosing manual over an automatic transmission, with other variables held constant.

**MODEL COMPARISON AND DIAGNOSTICS**

The adjusted R squared for this third model using stepwise backwards is `r round(summary(sw)$adj.r.squared,2)` which is higher than the adjusted R squared for the first model with only 1 factor, with `r round(summary(lm1)$adj.r.squared,2)`, so the third may represent a better model fit. Also see the Appendix for diagnostics with residuals plots of each model, and an ANOVA comparison.
 
\pagebreak 
  
##5. APPENDIX 

```{r summarystats,results='asis'}
# Summary statistics for the mpg
kable(t(as.matrix(summary(mtcars$mpg))), 
      caption = "Summary Statistics mpg",align = c("c", "c")) 
```
  
```{r plotviolin}
# Plot a violinplot to see the transmission types by mpg
mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am) <- c("automatic", "manual")
g <- ggplot(mtcars,aes(x=am,y=mpg))
g + geom_violin(aes(fill=am)) + 
  labs(title="Figure 1 - Violin plot of Transmission Type and Miles/(US) gallon", 
       x="Tranmission type",y="Miles/(US) gallon") + 
  scale_fill_discrete("Transmission Type")
```
  
**SIMPLE LINEAR REGRESSION MODEL**

```{r linearmodelsummary lm1}
##  Produce a summary of lm1 
summary(lm1)[4]
```
  
  
```{r residuals1}
## Model diagnostics Figure 2
par(mfrow=c(2,2))
plot(lm1)
```
  
1) Linear relationship between each (numerical) explanatory variables  
In the Simple linear regression there is only 1 categorical variable so we cannot check the residuals plot that residuals are scattered around 0.

2)  Nearly normal distribution of residuals   
The qqplot is almost a straight line we can say that this condition is met.

3) Constant variability of residuals  
Since these are randomly scattered around 0, this condition is met.

4) Leverage   
There are no patterns therefore this condition appears to be met

```{r dfbeta1 }
# Summary of dfbetas for model1
kable(dfbetas(lm1)[,2],caption ="Dfbetas for Model lm", 
      align = c("c", "c"))
```
  
From looking at there at the dfbetas, there do not appear to be any outliers or influence for the first model.

**MULTIVARIABLE LINEAR REGRESSION MODEL**

```{r MVRsummary}
## Produce a summary of  multivariable linear model using stepwise backwise model selection 
summary(sw)[4]
```
  
  
```{r residuals2}
## Model diagnostics Figure 3
par(mfrow=c(2,2))
plot(sw)
```
  
1) Linear relationship between each (numerical) explanatory variables  
These residuals are scattered around 0 so this condition is met.  

2)  Nearly normal distribution of residuals   
The qqplot is almost a straight line we can say that this condition is met.

3) Constant variability of residuals    
Since these are randomly scattered around 0, but this condition is met.

4) Leverage   
There are no patterns therefore this condition appears to be met. 
  
```{r varianceinflation}
# Take a look at the variance inflation factors 
kable(vif(sw),caption= "Table 5 - Variance Inflation factors Model sw",align = c("c", "c"))
```
   
   
We can see that the _cyl_, _disp_ and _wt_ have high variance in this model.
  
```{r dfbeta2 }
# Summary of dfbetas for sw model
kable(dfbetas(sw)[,2],caption="Table 6 - Dfbetas for sw model",align = c("c", "c"))
```
  
From looking at there at the dfbetas, there do not appear to be any outliers or influence for the third model.

Lastly we will compare the two models using anova.

```{r anova, resulst='asis'}
anova(lm1,sw)
```
Since the p-value is <0.05% we would reject a null hypothesis that the variable coefficients for model sw are 0 in favour of an alternate hypothesis that the coefficients are not 0.


**OTHER REGRESSION MODELS**

We would not use logistic regression since the _mpg_ outcome does not have two values but is a numerical outcome. Additionally we would not  use Poisson regression since the _mpg_ outcome is not a count.